---
title: "hw3_tpd2121"
output: github_document
date: "2024-10-12"
---

Initializing the packages
```{r initial}
library(tidyverse)
library(ggridges)
```

## Problem 2

2A. Loading the dataset
```{r dataset}
demography = read.csv("./nhanes_covar.csv", 
                      skip = 4) %>% 
  drop_na()
acc_data = read.csv("./nhanes_accel.csv")
```

Cleaning the dataset
```{r cleaning}
#checking data
anti_join(demography, acc_data)

#joining the dataset
complete_acc = 
  full_join(demography, acc_data, by = "SEQN") %>% 
  filter(age >21) %>% 
  mutate(
    sex = as.character(sex),
    sex = case_match(
      sex,
      "1" ~ "male",
      "2" ~ "female"), 
    education = as.character(education),
    education = case_match(
      education,
      "1" ~ "Less than highschool",
      "2" ~ "Highschool equivalent",
      "3" ~ "More than highschool"
    )
  )
```

2B. Creating table and visualization for gender, education, and age
```{r table}
complete_acc %>% 
  group_by(sex, education) %>% 
  summarize(count = n()) %>% 
  pivot_wider(names_from = "education", values_from = "count") %>% 
  knitr::kable()
```

```{r visualization}
complete_acc %>% 
  group_by(sex, education, age) %>% 
  ggplot(aes(x = education, y = age, color = sex)) + 
  geom_boxplot() + 
  labs(
    title = "Age distribution based on education and sex")
```

2C. Total activity over the day
```{r total}
total_df = 
  complete_acc %>% 
  mutate(
    act_total = rowSums(across(c(min1:min1440)))
    )%>% 
  select(SEQN:education, act_total)
```

```{r total_plot}
ggplot(total_df, aes(x = age, y = act_total, color = sex)) + 
  geom_point(alpha = .5) +
  geom_smooth(se = FALSE) + 
  facet_grid(. ~ education) + 
  theme_bw() +
  labs(
    title = "Total physical activity based on age, education, and sex",
    y = "Total physical activity per day")
#do facet_grid (sex ~ education) if want to make 6 panel
```

Based on the plot, we can see that there is a difference in the total activity based on age, education, and sex. For people who had education lower than highschool equivalent, the highest total activity per day is recorded at 30-50 years old group and decline afterwards. Female are shown to have higher activies per day compared to male in this group. People who have highschool equivalent education generally show decreasing trend of physical activity as they got older. This decline is more pronounced in female compared to male participants. There's a jump in activities recorded around 50 to 60 years of age, before declining again. Lastly, in the people with higher than highschool education, female participants' have plateauing activities until age 60 before declining. In male participants, there's a slight dip from around age 20s to 40s, before slightly jumping up and declining again at around age 50. 

2D. 3 panel plot
Accelerometer data allows the inspection activity over the course of the day. Make a three-panel plot that shows the 24-hour activity time courses for each education level and use color to indicate sex. Describe in words any patterns or conclusions you can make based on this graph; including smooth trends may help identify differences.
x axis = minutes, y axis = mean activity, color line = sex, panel = education

```{r time-course}
timecourse_df = 
  complete_acc %>% 
  select(-SEQN, -BMI, -age) %>% 
  group_by(sex, education) %>% 
  summarize(across(min1:min1440, mean)) %>% 
  pivot_longer(
    min1:min1440,
    names_to = "minutes",
    values_to = "activity"
    )
```

```{r timecourse_plot}
ggplot(timecourse_df, aes(x = minutes, y = activity, color = sex)) + 
  geom_point(alpha = .5) +
  geom_smooth(se = FALSE) + 
  facet_grid(education ~ .) + 
  theme_bw()

#pretty sure have to convert the minutes to hours, then plot the hours to the data
```

## Problem 3

3A. Load and clean the data
```{r load dataset}
jan_2020 = read.csv("./Jan 2020 Citi.csv") %>% 
  janitor::clean_names() %>%
  mutate(time = "jan_20") %>% 
  select(time, everything())

jan_2024 = read.csv("./Jan 2024 Citi.csv") %>% 
  janitor::clean_names() %>%
  mutate(time = "jan_24") %>% 
  select(time, everything())

jul_2020 = read.csv("./July 2020 Citi.csv") %>% 
  janitor::clean_names() %>%
  mutate(time = "jul_20") %>% 
  select(time, everything())

jul_2024 = read.csv("./July 2024 Citi.csv") %>% 
  janitor::clean_names() %>%
  mutate(time = "jul_24") %>% 
  select(time, everything())

citibike_df = 
  bind_rows(jan_2020, jan_2024, jul_2020, jul_2024)
```
There are 4 data set in the citibike folder. The January 2020 data contains `r nrow(jan_2020)`, while the January 2024 data contains `r nrow(jan_2024)` observations. Meanwhile, the July 2020 data contains `r nrow(jul_2020)` while the July 2024 contains `r nrow(jul_2024)` observations. The variables on all data frame are quite similar, which are `r names(jan_2020)`, with the variable `time` is an additional variable I created to help with merging the data into `citibike_df`. 

3B. Reader friendly table
```{r reader table}
citibike_df %>% 
  group_by(time, member_casual) %>% 
  summarize(count = n()) %>% 
  pivot_wider(names_from = "time", values_from = "count") %>% 
  knitr::kable()
```
From this table, we can see that in citibike members have higher ridership compared to casual riders from these 4 timepoints. We also see that while member ridership are increasing, the casual riders have a steep drop from `jul_20` to `jan_24`. 

3C. 5 popular stations
Make a table showing the 5 most popular starting stations for July 2024; include the number of rides originating from these stations.
```{r popular}
jul_2024 %>% 
  group_by(start_station_name) %>% 
  summarise(count = n())
  #rank function? and then make the plot
```

3D. Plot
Make a plot to investigate the effects of day of the week, month, and year on median ride duration. This plot can include one or more panels, but should facilitate comparison across all variables of interest. Comment on your observations from this plot.

```{r ride median}
citibike_df %>% 
  group_by(time, weekdays) %>% 
  summarise(
    median = median(duration, na.rm = TRUE)) %>% 
  ggplot(aes(x = weekdays, y = median, color = time)) + 
  geom_point(alpha = .5) +
  geom_smooth(se = FALSE) + 
  theme_bw()

#the trendline is missing
#x axis sorted alphabetically, not days in orders
```
From the plot, we can see that Jul 2020 has the highest ridership median compared to the other month-year timepoint. When looking at the days, we see that Saturdays generally have the highest median across all month-year time points. 

3E. Electric bikes
There were relatively few electric Citi Bikes in 2020, but many more are available now. For data in 2024, make a figure that shows the impact of month, membership status, and bike type on the distribution of ride duration. Comment on your results.

