---
title: "hw3_tpd2121"
output: github_document
date: "2024-10-12"
---

Initializing the packages
```{r initial, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggridges)
```

## Problem 2

**2A. Loading the dataset**
```{r dataset, message=FALSE, warning=FALSE}
demography = read.csv("./nhanes_covar.csv", 
                      skip = 4) %>% 
  drop_na()
acc_data = read.csv("./nhanes_accel.csv")
```

Cleaning the dataset
```{r cleaning, message=FALSE, warning=FALSE}
#checking data
anti_join(demography, acc_data)

#joining the dataset
complete_acc = 
  full_join(demography, acc_data, by = "SEQN") %>% 
  filter(age >= 21) %>% 
  mutate(
    sex = as.character(sex),
    sex = case_match(
      sex,
      "1" ~ "male",
      "2" ~ "female"), 
    education = as.character(education),
    education = case_match(
      education,
      "1" ~ "Less than highschool",
      "2" ~ "Highschool equivalent",
      "3" ~ "More than highschool"
    )
  )
```

**2B. Creating table and visualization for gender, education, and age**
```{r table, message=FALSE, warning=FALSE}
complete_acc %>% 
  group_by(sex, education) %>% 
  summarize(count = n()) %>% 
  pivot_wider(names_from = "education", values_from = "count") %>% 
  knitr::kable()
```

Based on the table, we see that there are slightly more male than female participants. Based on education we see that the highest number of participants have education more than high school and there are slightly more female than male in this category. The number of male and female with less than high school education was around the same, but there are more male compared to female in the high school equivalent category. 

```{r visualization, message=FALSE, warning=FALSE}
complete_acc %>% 
  group_by(sex, education, age) %>% 
  ggplot(aes(x = education, y = age, color = sex)) + 
  geom_boxplot() + 
  labs(
    title = "Age distribution based on education and sex")
```

I chose boxplot because it's able to show the distribution more clearly compared to histogram or density plot. Looking at the `more than high school` category, we see that despite female having slightly more participants in the category compared to male, we see that in terms of distribution, male have slightly older age compared to female in the category based on the interquartile range of the boxplot. For the `less than high school` category, the numbers and distribution are similar for both male and female, but we see that male has younger participants (longer bottom tail). And finally, for the `high school equivalent`, we see that despite female having lower number of participants, the interquartile range of the boxplot showed that female participants in the category is older compared to male. 

**2C. Total activity over the day**
```{r total, message=FALSE, warning=FALSE}
total_df = 
  complete_acc %>% 
  mutate(
    act_total = rowSums(across(c(min1:min1440)))
    )%>% 
  select(SEQN:education, act_total)
```

```{r total_plot, message=FALSE, warning=FALSE}
ggplot(total_df, aes(x = age, y = act_total, color = sex)) + 
  geom_point(alpha = .5) +
  geom_smooth(se = FALSE) + 
  facet_grid(. ~ education) + 
  theme_bw() +
  labs(
    title = "Total physical activity based on age, education, and sex",
    y = "Total physical activity per day", 
    caption = "Data from the NHAES package")
```

Using scatter plot, the data seems quite scattered and no pattern would seem clear for us readers if not for the smooth trend line added to each panel. Based on the plot, we can see that there is a difference in the total activity based on age, education, and sex. For people who had education `high school equivalent`, the highest total activity per day is recorded at 30-50 years old group and decline afterwards. Female are shown to have higher activities per day compared to male in this group. People who have `less than high school equivalent` education generally show decreasing trend of physical activity as they got older. This decline is more pronounced in female compared to male participants. There's a jump in activities recorded around 50 to 60 years of age, before declining again. Lastly, in the people with `more than high school` education, female participants' have plateauing activities until age 60 before declining. In male participants, there's a slight dip from around age 20s to 40s, before slightly jumping up and declining again at around age 50. 

**2D. 3 panel plot**
```{r time-course, message=FALSE, warning=FALSE}
timecourse_df = 
  complete_acc %>% 
  select(-SEQN, -BMI, -age) %>% 
  group_by(sex, education) %>% 
  summarize(across(min1:min1440, mean)) %>% 
  pivot_longer(
    min1:min1440,
    names_to = "minutes",
    names_prefix = "min",
    values_to = "activity"
    ) %>% 
  mutate(
    minutes = as.numeric(minutes),
    hours = minutes/60
    )
```

```{r timecourse_plot, message=FALSE, warning=FALSE}
ggplot(timecourse_df, aes(x = hours, y = activity, color = sex)) + 
  geom_smooth(se = FALSE) + 
  facet_grid(education ~ .) + 
  theme_bw() + 
  labs(
    title = "Progression of activity over 24h duration")
```

The line graph describe the progression of physical activity over 24 hours. The graph shows how physical activity corresponds to human general activity patterns and generally have similar patterns across education level. The observation start from hour 0, which is midnight, and we see declining trend between until hour 5, which corresponds to the early hours in the morning where people are going to bed/asleep. At around 5 AM, people tend to wake up so we see increasing activity until hour 20, or around 8 pm, when activities tend to slightly dip or plateau as people finished working. Female generally has higher activities compared to males, and people with less that high school education have the highest activities measurement compared to the other 2 education categories. 

## Problem 3

**3A. Load and clean the data**
```{r load dataset, message=FALSE, warning=FALSE}
jan_2020 = read.csv("./Jan 2020 Citi.csv") %>% 
  drop_na() %>% 
  janitor::clean_names() %>%
  mutate(time = "jan_20") %>% 
  select(time, everything())

jan_2024 = read.csv("./Jan 2024 Citi.csv") %>% 
  drop_na() %>% 
  janitor::clean_names() %>%
  mutate(time = "jan_24") %>% 
  select(time, everything())

jul_2020 = read.csv("./July 2020 Citi.csv") %>% 
  drop_na() %>% 
  janitor::clean_names() %>%
  mutate(time = "jul_20") %>% 
  select(time, everything())

jul_2024 = read.csv("./July 2024 Citi.csv") %>% 
  drop_na() %>% 
  janitor::clean_names() %>%
  mutate(time = "jul_24") %>% 
  select(time, everything())

citibike_df = 
  bind_rows(jan_2020, jan_2024, jul_2020, jul_2024)
```
There are 4 data set in the citibike folder. The January 2020 data contains `r nrow(jan_2020)` observations, while the January 2024 data contains `r nrow(jan_2024)` observations. Meanwhile, the July 2020 data contains `r nrow(jul_2020)` observations while the July 2024 contains `r nrow(jul_2024)` observations. The variables on all data frame are quite similar, which are `r names(jan_2020)`, with the variable `time` is an additional variable I created to help with merging the data into `citibike_df`. 

**3B. Reader friendly table**
```{r reader table, message=FALSE, warning=FALSE}
citibike_df %>% 
  group_by(time, member_casual) %>% 
  summarize(count = n()) %>% 
  pivot_wider(names_from = "time", values_from = "count") %>% 
  knitr::kable()
```
From this table, we can see that in citibike members have higher ridership compared to casual riders from these 4 time points. We also see that while member ridership are increasing, the casual riders have a steep drop from `jul_20` to `jan_24`. 

**3C. 5 popular stations**
```{r popular, message=FALSE, warning=FALSE}
jul_2024 %>% 
  group_by(start_station_name) %>% 
  count(start_station_name, name = "ride_count") %>% 
  arrange(desc(ride_count)) %>% 
  filter(ride_count >= 145) %>% 
  knitr::kable(
    col.names = c("Origin Station Name", "Number of Rides")
  )
  
```

**3D. Plot to investigate the effects of day of the week, month, and year on median ride duration**
```{r ride median, message=FALSE, warning=FALSE}
citibike_df %>% 
  group_by(time, weekdays) %>% 
  summarise(
    median = median(duration, na.rm = TRUE)) %>% 
  ggplot(aes(x = weekdays, y = median, color = time)) + 
  geom_point() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust  =1),
        legend.position = "bottom") +
  facet_grid(.~ time) +
  labs(
    title = "Ride duration over week and 4 timepoints", 
    x = "days of the week",
    y = "median of ride duration")
```

From the plot, we can see that `Jul_20` has the highest ridership median compared to the other month-year timepoint, while `Jan_24` has the lowest. Generally, Julys have higher ridership compared to Januarys. When looking at the days, we see that Saturdays and Sundays generally have the highest median across all month-year time points, except in Jan 24 where the highest median duration is on Friday and Saturday. 

**3E. Electric bikes**
```{r electric, message=FALSE, warning=FALSE}
citibike_df %>% 
  group_by(time, rideable_type, member_casual) %>%
  select(time, rideable_type, member_casual, duration) %>% 
  filter(time %in% c("jan_24", "jul_24")) %>% 
  ggplot(aes(x = duration, fill = rideable_type)) + 
  geom_histogram(position = "dodge", binwidth = 2) + 
  theme_bw() +
  facet_grid(member_casual ~ time) + 
  labs(
    title = "Electric bikes ridership based on types of riders and time", 
    x = "ride duration",
    y = "n number of rides", 
    color = "types of riders")
```

I used univariate plot, in this case histogram, since only 1 variable is numerical variable. Based on the plot, we can see that members always have higher ridership compared to casual riders and electric bikes have higher ridership compared to non-electric bikes. In terms of duration, Jul 2024 have longer ride duration for both members and casual riders compared to Jan 2024. 
